### 2014-07-25 Paper_v1.tex
Updated text and figures

\subsection*{Inhibition of Raf kinase activity modulated MAPK activation and differentiation markers.}
Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
For more details on GW5074 including dose response see Supplemental Materials including Fig. \ref{F:MAPKGWWB}A.
ATRA treatment alone statistically significantly increased the G1/G0 percentage over the untreated control, 
while GW5074 alone had a negligible effect on the cell cycle distribution (Fig. \ref{F:GWdiff}A).
Secondly, ATRA treatment alone increased both Raf phosphorylation at S621/S259 and Raf expression 24 hr post-treatment,
while ATRA-induced phosphorylation of ERK was blocked by GW5074, consistent with the loss of Raf kinase activity (Fig. \ref{F:MAPKGWWB}B).
Lastly, both the phosphorylation and expression of Raf was inhibited by GW5074.
This suggested a positive feedback between Raf kinase activity, Raf phosphorylation and ERK action on Raf transcription factors.
Surprisingly, the combination of GW5074 and ATRA statistically significantly increased the G1/G0 population (82 $\pm$ 1\%)
compared with ATRA alone (61 $\pm$ 0.5\%).
Increased G1/G0 arrest following the combined treatment with GW5074 and ATRA was unexpected, 
as the combination of ATRA and the MEK inhibitor (PD98059) has been shown previously to decrease ATRA-induced growth arrest \cite{Yen1998}.
However, growth arrest is not the sole indication of functional differentiation.
Expression of the cell surface marker CD11b has also been shown to coincide with HL-60 cells myeloid differentiation \cite{Hickstein1989}.  
We measured CD11b expression, for the various treatment groups, using immuno-fluorescence flow cytometry 48 hr post-treatment.
As with G1/G0 arrest, ATRA alone increased CD11b expression over the untreated control, 
while GW5074 further enhanced ATRA-induced CD11b expression (Fig. \ref{F:GWdiff}B).
GW5074 alone had no statistically significant effect on CD11b expression, 
compared with the untreated control.

>>>>

\subsection*{Mathematical modeling of ATRA-induced MAPK activation in HL-60 cells.}
We constructed a mechanistic mathematical model of a putative Raf-Vav1 activation circuit (Fig. \ref{F:network}).  
The immunoprecipitation and kinase inhibitor studies suggested that Raf kinase activity was required for ATRA-induced ERK phosphorylation and ROS response. 
Furthermore, the phosphorylation, expression, and Vav1 binding of Raf itself decreased in response to Raf inhibition. 
These results suggested a possible positive feedback loop between Raf kinase activity and increased Raf expression, 
possibly through ppERK-dependent transcription factors such as Sp1 and STAT1 \cite{Kim2009,Zhang2004}.
Both Sp1 and STAT1 have putative binding domains in the Raf promoter (TRANSFAC).
However, this loop is not ATRA-responsive, without an ATRA-inducible trigger, which we hypothesize is Vav1.
The Vav1 gene contains an upstream canonical Retinoic Acid Response Element (RARE) sequence motif (TRANSFAC) which could allow
Vav1 expression to be at least partially regulated by ATRA. 
By contrast, the Src promoter does not have a discernible canonical putative RARE.
(Tables \ref{T:params}, \ref{T:species} and \ref{T:rxns} respectively).
The levels of ATRA and GW5074 were introduced as user defined parameters depending on the experiment being simulated.
In general Dilution due to growth was 
More details on the model architecture including additional references can be found in Materials and Methods.

To address parametric uncertainty, 
an ensemble of models was estimated using the eight datasets taken from this study (38 total data points),
and the previous study of Wang and Yen \cite{Wang2008}.
Here ensemble refers to the use of different parametric sets for the same system of reactions.
Cross-validation (data randomly split into 80\% training and 20\% validation) demonstrated that, 
on average, the model identification scheme (described in Materials and Methods) decreased both training 
and prediction error (Fig. \ref{F:parameters}) over successive iterations of the parameter selection procedure. 
This is suggestive of the predictive ability of the model and also demonstrated that the nominal parameter set, 
the set with the lowest training error, did not result in over-fitting. 
Subsequently, we considered an ensemble of likely parameter sets (N = 99) with relative mean squared training errors within 20\% of the nominal set. 
The model ensemble recapitulated both time-series data (including sustained MAPK activation) and the GW5074 results (Fig. \ref{F:Train}).

\subsection*{Inhibition of Raf kinase activity modulated MAPK activation and differentiation markers.}
We considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
For more details on GW5074 including dose response see Supplemental Materials including Fig. \ref{F:MAPKGWWB}A.
ATRA treatment alone statistically significantly increased the G1/G0 percentage over the untreated control, 
while GW5074 alone had a negligible effect on the cell cycle distribution (Fig. \ref{F:GWdiff}A).
Secondly, ATRA treatment alone increased both Raf phosphorylation at S621/S259 and Raf expression 24 hr post-treatment,
while ATRA-induced phosphorylation of ERK was blocked by GW5074, consistent with the loss of Raf kinase activity (Fig. \ref{F:MAPKGWWB}B).
Lastly, both the phosphorylation and expression of Raf was inhibited by GW5074.
This suggested a positive feedback between Raf kinase activity, Raf phosphorylation and possible ERK action on Raf transcription factors.
Surprisingly, the combination of GW5074 and ATRA statistically significantly increased the G1/G0 population (82 $\pm$ 1\%)
compared with ATRA alone (61 $\pm$ 0.5\%).
Increased G1/G0 arrest following the combined treatment with GW5074 and ATRA was unexpected, 
as the combination of ATRA and the MEK inhibitor (PD98059) has been shown previously to decrease ATRA-induced growth arrest \cite{Yen1998}.
However, growth arrest is not the sole indication of functional differentiation.
Expression of the cell surface marker CD11b has also been shown to coincide with HL-60 cells myeloid differentiation \cite{Hickstein1989}.  
We measured CD11b expression, for the various treatment groups, using immuno-fluorescence flow cytometry 48 hr post-treatment.
As with G1/G0 arrest, ATRA alone increased CD11b expression over the untreated control, 
while GW5074 further enhanced ATRA-induced CD11b expression (Fig. \ref{F:GWdiff}B).
GW5074 alone had no statistically significant effect on CD11b expression, 
compared with the untreated control.
